taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-07-21 paper/paper.Rmd
:pen:

The National Ecological Observatory Network (NEON) provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.  `neonstore` design emphasizes simplicity and data provenance. here, we describe the design, use, and rationale for the package.

NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.  Products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, (particularly relevant to R users where most objects are typically stored in memory)

## Installation

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/neonstore")
```
## Quickstart

Discover data products of interest:

```{r}
products <- neon_products()

i <- grepl("Populations", products$themes)
products[i, c("productCode", "productName")]

i <- grepl("bird", products$keywords)
products[i, c("productCode", "productName")]
```

Download all data files in the bird survey data products.

```{r}
library(neonstore)
neon_download("DP1.10003.001")

```

@@

>>>>

The National Ecological Observatory Network (NEON) provides open access ecological data as 181 (active and planned) data products at 81 (permanent and relocatable) sites throughout the United States.  The `neonstore` R package seeks to provide quick, convenient access and persistent storage of NEON data files.  `neonstore` design emphasizes simplicity and data provenance. Here, we describe the design, use, and rationale for the package.

## Rationale and package design

NEON data products are built around a highly atomized file design: both observation systems (OS) and instrument systems (IS) data products are divided into separate files by site and month.  Meanwhile, products generated by the airborne observation platform (AOP) are typically divided by type, flight date, flight line, and grid cell.  This atomized approach facilitates sub-setting and data transfer, because users can identify ahead of time which products, locations, and time ranges they need. As some NEON data products can be quite large (NEON's data catalog to date is in excess of 1 petabyte, primarily due to the size of many AOP products), being able to only download or only read into R the files of interest is already important to save bandwidth or computer memory, and this will be increasingly true as NEON data accumulates in the years ahead. This approach also facilitates data provenance and reproducibility, because it means that an update or correction to data collected at a particular site and time results only in a change to a single file and not the whole data product. Users can easily verify whether or not the update has changed data in their analyses by comparing file hashes, rather than having to read in and examine data meticulously row by row.

However, the highly atomized structure of NEON's data can also become a barrier to entry without convenient tools for working
with many files.  Most analyses will inevitably require multiple files to be combined or stacked, e.g. to create a unified record for a product across many months or years of sampling or across many sites.  When the process of combining files is complex or slow to execute, many researchers will resort to saving the resulting 'stacked' products as a basis for future work.  Such workflows may be more cumbersome to update as additional data continues to be published, or to compare against any subsequent corrections to previously-published data.  `neonstore` seeks to help users download and manage NEON files locally in a way that
lets analyses be quickly traced back to the raw data files.

`neonstore` is distinguished as much by what it does not do as by what it does.  All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface. Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.  A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.  Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools. When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.

## `neonstore` Functionality

`neonstore` functionality can be divided into two halves: functions that interface with the NEON Application Programming Interface (API), and functions for working with a local store of previously-downloaded NEON files.

```{r}
library(neonstore)
```

### NEON API functions

`neon_products()` and `neon_sites()` bind the NEON REpresentational State Transfer (REST) API endpoints for `/products` and `/sites`, providing information about each as a data frame. This can be useful for product discovery and occassionally for referencing other useful metadata about products and sites.  Each of these functions requires only a single API call (see discussion of rate limiting below).

```{r}
products <- neon_products()
products
```

```{r}
sites <- neon_sites()
sites
```

The workhorse function of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.  For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`:

```{r}
neon_download("DP1.10003.001")
```

To download this data, `neon_download()` must make many API calls.  It first uses `neon_sites()` to reference all sites containing this data product.  Then for each site, for each month for which that site collected data, the function will make an API call to the `/data` endpoint (using the `neonstore` function, `neon_data()`), which returns a table of files found for that product at the requested site and month range.  The  table from `neon_data()` indicates 4 values: the file name, the MD5 hash of the data file (confusingly labeled as the crc32 sum, though CRC32 hashes are used only for AOP data products.  Note that MD5 hashes are length-32 character strings, while the CRC32 hashes are 8 characters long), the file size (in bytes -- encoded as a character because some file sizes are too long to be represented as an integer), and a download URL.  The download URL embeds a temporary access token which expires after an hour, so users must complete all the downloads before this time expires.  The access string for each file is unique, thus it is (at the time of writing) impossible to download any given data file without making a corresponding API request, even if the precise file name and address is already known.

The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling.

@@

Note that at this time, `neonstore` does not provide a complete
